---
title: "Homework 3"
author: ""
date: "2024-09-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Read the data in the chunk below. The data come from of a study that investigated the flood tolerance of three different clover species.  
The data can be interpreted as arising from an experiment with a 3 $\times$ 2 factorial treatment structure in a completely randomized design with 24 repetitions. The experimental units were pots in the greenhouse. The researchers measured the crown biomass, which represents the amount of storage carbohidrates in a plant and can indicate plant stress (i.e., lower biomass is associated to more stress).    

```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

```{r }
url <- "https://raw.githubusercontent.com/jlacasa/stat705_fall2024/main/classes/data/lotus_part4.csv"
dd <- read.csv(url)
```

```{r}
dd %>% 
  ggplot(aes(species, crown_g))+
  geom_boxplot(aes(fill = trt))+
  scale_fill_manual(values = c("#43AA8B", "#DB504A"))+
  labs(x = "Species", 
       y = "Crown biomass (g)", 
       fill = "Species")+
  theme_classic()+
  theme(aspect.ratio = 1)
```

### a. Propose a statistical model (using mathematical notation) to describe the relationship between crown biomass, legume species, and fit that model to the data.  

### b. Mention the assumptions in your model and check them. 