---
title: "Day 6 - 08/30/2024"
format: html
editor: visual
---

## From last class
- **Assignments**. A few comments:  
  - Good overall.  
  - `geom_smooth()` is forbidden from now on! Use `stat_function()` instead.  
  - How is a statistical model built.
    - Deterministic function
    - Likelihood function
    - Method of estimation
  
## Linear models revisited  

Let's look at the clover data again. 

```{r}
url <- "https://raw.githubusercontent.com/jlacasa/stat705_fall2024/main/classes/data/lotus_hw1.csv"
data <- read.csv(url)
```

```{r include=FALSE, eval=TRUE}
library(tidyverse)
```

```{r}
data %>% 
  ggplot(aes(doy, stm.length_cm))+
  geom_point()+
  labs(x = "Day of the year", 
       y = "Stem length (cm)")+
  theme_classic()+
  theme(aspect.ratio = 1)
```

What's a good model for the data?  

One option is 
$$y_i = \beta_0 + x_i \beta_1 + \varepsilon_i, $$
$$\varepsilon_i \sim N(0, \sigma^2),$$

for $i = 1, 2, ..., n$ (n being the total number of observations), where $y_i$ is the length of the stem (in cm) of the $i$th observation, $\beta_0$ is the length of the stem (in cm) at day of the year 0 (i.e., December 31 of last year), $x_i$ is the day of the year of the $i$th observation, $\varepsilon_i$ is the residual of the $i$th observation, that is normally distributed. 

This is the same as writing  

$$\mathbf{y} \sim \text{N}(\boldsymbol{\mu}, \sigma^2\mathbf{I}),$$
$$\boldsymbol{\mu}=\mathbf{X}\boldsymbol{\beta}.$$

### Estimation  

Let's fit the model written above to data. [R script](https://github.com/jlacasa/stat705_fall2024/blob/main/classes/in_class_code/day6_estimation_example.Rmd)

```{r include=FALSE, eval=TRUE}
m_cont <- lm(stm.length_cm ~ 1 + doy,  data = data)
```

#### Confidence intervals  

```{r}
confint(m_cont)
```


### How much money would you bet on the estimate? Let's do a simulation.  
R code [here](https://github.com/jlacasa/stat705_fall2024/blob/main/classes/in_class_code/day5_simulation_example.qmd).  

## For next class  

- No class on Monday.  
- Review your assignments.

